Topic 6.2: Open Data

### 2020-05-05 draft_chapter.Rmd
Edits and additions 2

---
title: "Topic 6.2: Open Data"
author: "Samuel Langton (MMU) & Reka Solymosi (University of Manchester)"
date: "June 2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

## Introduction

- Topic definition.
- Advantages and exciting prospects in open data.
- Chapter structure outline.
- Prerequisites (reference _Geocomputation with R_).

## Background

- What is open data?
- What types of organisations release open data?
  - Public sector
  - Private sector
- Why is it needed?
- What are the strengths?
- What are the limitations?
- Substantive examples in criminology (focus on crime and place research).
- Conclude with: a key one is Open Street Map.

## Open Street Map

### What is Open Street Map?

- Brief history.
- Motivations for and advantages of an open web mapping platform.
- Primary features (keys, values, elements).
- Contributing to the data yourself.

### Downloading Open Street Map Data

- Define and explain APIs.
- For OSM, describe overpass queries and overpass-turbo.
- Query example (brief).

### Using Open Street Map data in R

- Made easy through the `osmdata` package in R.
- More straightforward than overpass-turbo.
- Compatible with both `sp` and `sf` classes of spatial data.
- We focus on `sf` as its compatible with the `tidyverse`.
- Outline the key documentation and references.

@@

First, ensure that you have the relevant packages installed in R. Although the main package used in this demonstration is `osmdata`, for querying the Open Street Map API, we use a number of additional packages for data handling and visualisation. If you don't have these packages installed, use the `install.packages()` function prior to loading each one with `library()`.

```{r, eval = F}
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

>>>>

---
title: "Topic 6.2: Open Data"
author: "Samuel Langton (MMU) & Reka Solymosi (University of Manchester)"
date: "June 2020"
output:
  pdf_document:
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
library(jsonlite)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

# Introduction

Access to data in crime and place research, and even criminology more broadly, has traditionally been reserved to those who have the means to collect fresh data themselves, pay for access, or obtain data through formal data sharing agreements. Even when access is granted, the usage of these data often comes with conditions that circumscribe how the data can be used through licensing or policy [@kitchin2014data]. The public dissemination of findings which emerge from analysis might even be subject to restrictions. This can lead to unequal access, controlled usage and curb the diffusion of findings, severely limiting the insight that can be attained from data.

Open data are a response to these shortcomings, broadening access and participation in research, removing the requirement for permissions, formal agreements and negotiations [@manovich2011trending]. Not only this, but open data can also lead to all sorts of novel insight within criminology and cime and place research, tapping into constructs and processes which are difficult to capture through surveys, interviews and other traditional meaures [@solymosi2018role]. As such, it is important that social scientists, researchers, crime analysts, and others interested in making sense of the social world around them have the skills and know-how to access, accurately interpret, assess and analyse open data sets.

This chapter aims to lay the groundwork for developing such skills by providing a framework to approach and meaningfully interpret open data, in terms of understanding its origin and the implications this might have on the data itself. The chapter also offers a practical hands-on guide to accessing, wrangling, and analysing open data in order to draw conclusions about crime and place. We achieve this by first giving a background on open data, discussing the key types of data out there, critically assessing the strengths and limitations, and building a how-to guide checklist for researchers to follow when approaching an open data source. Finally, we work through an example that shows how to access, wrangle, link, and interpret open data from various sources, providing a template which can be applied in other research areas.

# Background

## What is open data?

First, it is useful to clarify what we mean when we refer to 'open data'. The _Open Data Handbook_, compiled by the Open Knowledge Foundation, states that open data are "data that can be freely used, re-used and redistributed by anyone - subject only, at most, to the requirement to attribute and sharealike" [@dietrich2009open]. Specifically, open data can be defined along three key domains [@knowledge2016open].

- **Availability and Access:** the data must be accessible via a public domain, at no more than a reasonable reproduction cost, and through a convenient medium, such as the internet. Ideally, it should be machine-readable and modifiable. For example, it should be downloadable from a website as an unlocked spreadsheet, or json file, rather than presented as a summary table in a PDF document.
- **Re-use and Redistribution:** the data must be provided under conditions that allow free use, such as modification, separation, or compilation of the data and permit re-use and redistribution including the intermixing (merging) with other datasets.
- **Universal Participation:** everyone should be allowed to use, re-use and redistribute the data and its derivatives without restriction. For instance, restrictions that only allow non-commercial or educational usage would not constitute open data.

In short, 'open data' should be data that are readily and publicly available, in a usable format, and without reistrictions on usage, modification and re-distribution.

Open data practices, and the extent to which it is a reality, vary both within and between countries. For example, the United States have a history of making public sector data sets openly available. The United Kingdom can be more strict, releasing data under a licence ('Open Government Licence'), with some data requiring a fee [@kitchin2014data]. The _Global Open Data Index_ [@globopendata] is one tool for tracking the annual global benchmark for publication of open government data by countries. It providing a rank which indicates how well governments across the world follow open data practices on aggregate, but also according to specific domains. Many of these have direct relevance to crime and place research, such as administrative boundaries and geographic locations (e.g. coordinates).

## What are types of open data?

Now we have a reasonable idea about what open data is in a broad sense, we can consider the different types of open data which you might come across. Here, we formulate a typology of open data sets based on their origin to help provide an overview.

### Public sector (including publicly funded research).

Much of the open data movement has focused on information generated by local and national state agencies, or publicly funded research. Open data has become a key component of transparency and accountability in government, given that much of the data collected has been funded by the public purse [@kitchin2014data]. In some countries, the national statistics body is independent from government, which serves to improve accountability and increase public trust in the data being collated and published. The _Global Open Data Index_ introduced earlier focuses on identifying the gaps in governmental organisations, encouraging them think about how public sector information can become more usable, and ultimately, more impactful [@globopendata].

As researchers studying crime and place, public sector data is particularly important. Many variables thought to explain the non-random distribution of crime, offenders and victims in space are coallated and published by governmental bodies. For instance, demographic characteristics about resident populations can be collected via national censuses, with data usually released at aggregations which constitute 'neighbourhoods' - a theoretically important scale in criminology. Data describing urban structure, such as street networks, are often generated and distributed by public bodies, and can be used in street syntax analysis to assess whether the configeration of streets dictates things like violent crime victimisation [@streetsyntax2017]. In many countries, victimisation surveys are openly available for download. This is not only useful for substantive research (e.g. quantifying long-term crime trends), but also scrutiny over reliability and accuracy in other forms of open data, such as police recorded crime records. Both of these examples represent an interesting challenge to open data principles, since data tends to be anonymised to some degree before being made available to the public. Access to the raw records is dependent on a number of requrements (e.g. background checks, contractual agreements, ICT security measures) which might exclude many individuals due to cost and practicality.

The drive towards transparency and scrutiny of public information is further enhanced by an increasing need for replication - a hallmark for open science research practice. Opening up data sets used in criminology publications and wider social sciences fosters and facilitates a culture of replication [@pridemore2018replication]. As such, open data fits into the wider discourse around transparency, open review and open scrutiny, which (in time) may even become a policy requirement irrespective of the research field or the source of funding [@vuong2017open]. Increasingly, researchers are making use of online repositories for their papers (e.g. www.osf.io), data (e.g. www.ukdataservice.ac.uk/deposit-data) and code (e.g. www.github.com), all in the name of transparency and replication. These are guidelines and tools worth considering, not just when conducting analysis on secondary data sources, but also collecting and sharing your own data.

### Private sector

Opening up data generated by the private sector represents a significant challenge, largely due to the proprierary value to its creators [@kitchin2014data]. The aims and objectives of private companies, who have a duty to shareholders and operate in a competitive market environment, differ considerably from local and central government. That said, some datasets are released openly by private sector organisations, albiet often only a subset of what would be available as a paying customer. The website for ArcGIS, a piece of software maintained by the Environmental Systems Research Institute (ESRI), a private company, host a number of open geographic datasets on their online cloud platform (www.hub.arcgis.com/search). Indeed, many papers have made use of data collected or distributed by private organisations to explore crime and place, such as Google Street View [@langton2017residential] and Twitter [@malleson2015impact]. In this book you will learn to use the free Twitter API for such purposes in criminology (see Section 3, Topic 7).

### Open crowdsourced data

Finally, there is a specific group of open data sources that collate data collectively, generated by large groups of individuals who do not specifically belong to any organisation, but instead work together on a collaborative project. These are crowdsourced data sets. Examples include Wikipedia (www.wikipedia.org), an online encyclopedia where anyone can contribute, and Flickr (www.flickr.com) an online photo gallery where people can upload and tag their photos. Often, the organisations who maintain and monitor these data collection activities are charities or non-govermental organisations that do not operate for profit, but instead provide some form of social good. An example is the online reporting platform FixMyStreet (www.fixmystreet.com), where people can report problems such as instances of graffiti, vandalism or environment issues. The data has been used to explore signal crimes theory, for instance, offering insight into people's experiences with incivilities [@solymosi2018crowdsourcing]. Another chapter in this book (see Section 3, Topic 6) discusses the merits and pitfalls of crowdsourced data, and what to watch out for when analysing open data of this type.

## Strengths and limitations

The defining characteristics of open data, namely, that of availability, re-usability and universal participation, outlined earlier, represent its greatest strength. But besides from this, there are other advantages. One specific motivation for using open data comes from its potential to address many of the limitations associated with traditional surveying methods, such as social desirability bias, or issues associated with memory and recall [@mayer2013big]. This advantage exists largely due to the organic way in which open data is generated. It is often a by-product of other activities, and as such, we can gain an honest insight into people's everyday lives and associated social processes [@solymosi2018role]. As noted, open data also means open research, facilitating transparency and reproducibility. That said, there are a number of shortcomings which are worthy of consideration.

Firstly, one of the biggest obstacles to collecting open data is the ability to interpret what the data means (e.g. applying a framework for its analysis), and computational skills in data scraping, wrangling and cleaning, in order to transform it into a usable format for research [@boyd2012critical]. All sorts of open data remain unanalysed, and or inaccessible, to large portions of the population, simply because they lack the skills and know-how. Although this is certainly an obstacle for data usage more generally, it represents a key challenge to the governmental bodies, in particular, on which there is an onus to ensure transparency and faciliate scrutiny. Private companies, especially those who generate open data as a by-product of their primary activity (e.g. Twitter), have little responsibility or pressure to ensure that their data is accessible and usable to the general public. Moreover, in both public and private spheres, licences and conditions on open data are not necessarily concrete, and might be subject to change with little or no notice. This is an important consideration when planning and running long-term research projects.

Secondly, despite the merits of open data in terms of remedying pitfalls in traditional survey methods, there are new threats to validity that may emerge. For instance, whilst many online resources offer an 'honest' depiction of society, uncaptured by surveys, researchers should be careful in interpreting people's communication online as completely authentic [@manovich2011trending]. This could be a result of individuals willingly managing and 'curating' their online presence, or because of wider issues such as government consorship, or cultural norms around particular topics such as sexual assault. Relatedly, researchers should be aware of issues over sampling, and ultimately the generalisability of findings that emerge from the analysis of open data. In the case of crowdsourced data, the sample is not randomly drawn from a population, but rather, it is self-selected, giving way for people willing to discuss or contribute to a particular issue, which introduced a degree of bias [@longley2012geodemographics]. Specifically, contributors tend to be men, between the ages of 20-50, with a college or university degree [@budhathoki2010participants; @haklay2010good]. Contributions to resources such as Open Street Map, which we look at in the practical exercise later, are correlated with contextual characterisics such as poverty and population density, and as such, coverage is non-uniformly distributed across urban areas [@mashhadi2013putting]. These are important to keep in mind (and be transparent about) when reporting findings based on analysis of such data.

Open data can be messy, biased and noisey, but criminology (and social sciences more generally) can benefit immeasurably from its use. Only through open data can public sector bodies be held to account, research be transparent and reproducible, and participation in data analysis universal. In crime and place research, both dependent variables (e.g. police recorded crime incidents, victimisation rates) and independent variables (e.g. demographic characteristics, ambient population estimates) can be sourced from open data, whether public, private or crowdsourced. With a critical, engaged and considered approach to conducting research with open data, criminology can become a leading force in open and reproducible social science. In sum, researchers and analysts using open data must first ask critical questions:

- **Where** does this data come from? 
- **Why** was the data collected in the first place?
- **Who** is represented in the data, and who is excluded?
- **What** concepts and constructs can and cannot be operationalised with this data?
- **When** did data collection take place, and how might have that influenced results?

With this in mind, the following exercise utilises two different sources of open data on public transport and police recorded crime in London, England.

# Practical exercise

In this exercise we will demonstrate a number of different skills, including:

- Accessing open data using Application Programming Interfaces (APIs) and direct downloads.
- Cleaning and wrangling data for analysis.
- Visualising open spatial data.
- Comparing different sources of open data.
- Engaging with the critical questions outline above to better understand open data.

## Our aim

In demonstrating these skills, we will be answering two key research questions. The first is a substantive empirical investigation into crime in and around London Underground stations, in the spirit of existing research into transport crime (see newton2014above). Studies in Sweden, for instance, have shown the importance of environmental and neighbourhood characteristics in determining crime concentrations at underground stations in Stockholm, along with the positioning of stations on the line [@ceccato2013security]. Investigations have also taken place into crime concentrations at bus stops in Los Angeles [@loukaitou1999hot] and the impact of intensive policing along bus corridors in Merseyside, England [@newton2004crime]. As auch, we can pose the following research question:

- RQ1: To what extent does crime cluster in and around London underground stations?

The second is more methodological, and relates to our considered and critical usage of open data, by exploring the validity of different open data sources. As such, we can pose:

- RQ2: How comparable are public sector and crowdsourced open data on London underground stations?

First, we will go through how to access the data used in the exercises: an important skill when seeking to use open data.

## Accessing data

We will be using two different types of open data in this exercise, public sector data (crime data and transport authority data) and crowdsourced data (from Open Street Map). To access them we will use three different methods: 1) Direct download, 2) Direct request to an API, 3) Request to an API using a wrapper. This will give you few different ideas about how you might go about accessing other open data sets relevant to your research.

### Direct download

The simplest way that open data is often made available is through direct download from a website. This is the case for open police data made available in the United Kingdom through the [data.police.uk](https://data.police.uk/) webportal. To access this data, simply type the url into any web browser ( [https://data.police.uk/](https://data.police.uk/) ), and follow the guidance to acquire data. In the case of police.uk you [select the tab which says "Data"](https://data.police.uk/data/), choose your "Date range:" (ie for what months you would like to download crime data for), choose the police forces you want crime data from, choose some options (whether you want to include crime data, outcomes data, and/or stop and search data) and click on 'Generate File' to download.

Once you download this data you will have to save this locally on your computer. We have saved ours in a subfolder called "data" and will be loading the data into R directly from there.

Some data downloads may come bundled with a data dictionary, explaining important information about your data set, which you will need to answer the critical questions outlined in previous sections. In other cases this might be found online. This is the case with the police.uk dataset, where the data dictionary can be found on the ["About" page](https://data.police.uk/about/). Make sure you always read through this correctly so you can answer your critical questions and fully understand your data set.

### Direct request to an API

An application programming interface (API) is a tool which defines an interface for a programme to interact with a software component or system, for example, what sort of requests or calls can be made to it, and how these can be made. We are specifically using the term APIs to denote tools created by an open data provider to provide access to different subsets of their content. APIs facilitate scripted and programmatic extraction of content, as permitted by the API provider [@olmedilla2016harvesting]. APIs can take many different forms and be of varying quality and usefulness [@foster2016big]. For the purposes of accessing open data from the web, we are specifically talking about **RESTful APIs**. REST stands for Representational State Transfer. These APIs work directly over the web, meaning that we as users can play with the API with relative ease to understand how it works [@foster2016big].

Here we will use the example of the British transit authority Transport for London (TfL) to practice constructing HTTP requests via their API, and then parsing the data that are returned. Much like how downloading data directly requires access to a data dictionary to make sense of the data and answer our critical questions, we must find a similar document to understand how the API we plan on using works. In the case of the TfL API this is provided via the [TfL Open Data page for their unified API](https://tfl.gov.uk/info-for/open-data-users/unified-api) and the [documentation page](https://api-portal.tfl.gov.uk/docs). We will engage more with this in the exercises.

### Request to an API using a wrapper

Often developers who work with APIs will share their code for doing so, and release them in the form of a package or module. This is called a **wrapper** because it's code that wraps the API in a nice package, making it easier for future users to access. The wrapper can take many forms, it could be a Python module, or an R package, or it can even be a web interface that provides a graphical user interface (GUI) for accessing the API in question.

To demonstrate this, we will be accessing data from Open Street Map (OSM). OpenStreetMap is a crowdsourced open data set of spatial information built by a community of mappers that contribute and maintain data about all sorts of environmental features, such as roads, trails, cafes, railway stations, and much more, all over the world. The result is a map rich in local knowledge, which prvides valuable information about places and their features.

OSM has two types of wrappers available for its API, a web-based GUI called [Overpass Turbo](https://overpass-turbo.eu/), and an R package called `osmdata`. We will explore using both.

<!-- ### Open Street Map -->

<!-- #### What is Open Street Map? -->

<!-- - Brief history. -->
<!-- - Motivations for and advantages of an open web mapping platform. -->
<!-- - Primary features (keys, values, elements). -->
<!-- - Contributing to the data yourself. -->

<!-- ### data.police.uk -->

<!-- ### Transport for London Open Data Portal -->

<!-- ### Downloading Open Street Map Data -->

<!-- - Define and explain APIs. -->
<!-- - For OSM, describe overpass queries and overpass-turbo. -->
<!-- - Query example (brief). -->

<!-- ### Using Open Street Map data in R -->

<!-- - Made easy through the `osmdata` package in R. -->
<!-- - More straightforward than overpass-turbo. -->
<!-- - Compatible with both `sp` and `sf` classes of spatial data. -->
<!-- - We focus on `sf` as its compatible with the `tidyverse`. -->
<!-- - Outline the key documentation and references. -->

@@

Let's get started! Getting back to our problem at hand, crime at bus stops, we want to know how crimes and bus stops interact...
<!-- clear up what research q we can actually answer....!! -->

First, ensure that you have the relevant packages installed in R. Although the main packages used in this demonstration are `rjson` for parsing the TfL API results and `osmdata` for querying the Open Street Map API, we use a number of additional packages for data handling and visualisation. If you don't have these packages installed, use the `install.packages()` function prior to loading each one with `library()`.

```{r}
library(jsonlite)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

Now let's focus on acquiring our different data sets

### Direct Dowload of crime data

<!-- IDK how much detail we actually need here, probably not loads, about downloading extracting and importing police.uk data -->

### Direct request to TfL API

Our first step is to read through the . The tells us what sort of data is possible to acquire. We know that currently we want to find the location of bus stops, and any information that is available from TfL about these bus stops. In the documentation is where we find specifically *how* we can request this.

Specifically, we can scroll down to where we find "". Next to this we see the example URL: `https://api.tfl.gov.uk/line/24/route/sequence/outbound` This example demonstrates how we can request this information, and gives us a template to alter, but also allows us to take a peak into the result this call will return.

To see, try to take the above URL, and copy and paste it into your browser window. What do you see? It should be something like this:

```{r tfl, out.width = "\\textwidth", fig.cap="Data from TfL"}
knitr::include_graphics("img/tfl_screenshot.png")
```

This contains the information we requested (bus route 24 stops outbound) in a format called **JSON**, which stands for JavaScript Object Notation. JSON is an open standard way of storing and exchanging data, and will most likely be the format in which data are returned from most API calls.

Great, but how do we get this into R? Well this is where the `rjson` package is handy, and specifically we can use the `fromJSON()` function to parse this JSON p=object, which we can read directly from the URL into R using the `readLines()` function (base R). So, if we wanted the results of the bus route 24, we would type:

```{r getrt24, eval = F}
api_call <- fromJSON(readLines("https://api.tfl.gov.uk/line/24/route/sequence/outbound"))
```

This gives us an object (`api_call`) which contains all the information returned by the TfL API. Now JSON is slightly different to dataframes which we are more used to dealing with, in that there is nested information

```{r conferttodf, eval = F}
json_data_frame <- as.data.frame(api_call)
```

<!-- insert here tutorial on parsing the api data, this is a job for next week i'm tired -->

### Using wrappers to access Open Street Map data

<!-- Maybe start with the GUI-based one, and then do the below in R? -->
